Supporting Information : The importance of a complete statistical description of dependence between variables for ecology and related fields

### 2018-06-26 Paper.Rmd
adding response by **SG

@@

Left- and right-tail dependence quantify the extent to which two ranked
variables $u_i$, $v_i$ for $i=1,...,n$ 
(see \nameref{Model_selection} for defintions of $u_i$ and $v_i$)
are related to each other in the left and right tails, respectively, of their 
distributions [@nelsen2007_copula, @joe2014_dependence]. 
Our nonparametric study of tail dependence uses three
statistics which can quantify the extent to which $u_i$ and $v_i$ are 
related in any part of their distributions. We here describe the 
statistics. The statistics were formulated with positively associated
variables in mind. All the variables we study are positively associated
when they are significantly associated. So definitions and associated 
diagrams will assume positive associations, though generalizations are 
possible. Given two bounds 
$0 \leq l_b < u_b \leq 1$, we define the parallel lines $u+v=2l_b$ and 
$u+v=2u_b$ which intersect the unit square in which the $u_i$ and $v_i$ 
fall (these are the blue and red lines, respectively, in 
Fig. \ref{fig_stat_image}(A)). Our statistics quantify the dependence
between $u_i$ and $v_i$ in the region bounded by these lines.
Using $l_b=0$ and some value $u_b\leq 0.5$ can give information about 
dependence in the left parts of the distributions of $u$ and $v$, and 
using $u_b=1$ and $l_b \geq 0.5$ can give information about dependence in 
the right parts of the distributions.

\begin{figure}[!h]
\textbf{ \hspace{6 cm} (A) \hspace{5.6 cm} (B)} \\
\vspace{-0.6 cm}
\begin{center}
\hspace{0.5 cm}
\includegraphics[width=5 cm,height=4.3 cm]{./stat_image.jpg} \hspace{0.6 cm}
\includegraphics[width=6 cm]{./Results/stat_results/stat_testing/CorlmCoru_C.pdf}
\end{center}
\textbf{ \hspace{6 cm} (C) \hspace{5.7 cm} (D)} \\
\vspace{-0.6 cm}
\begin{center}
\hspace{0.2 cm}
\includegraphics[width=6 cm]{./Results/stat_results/stat_testing/PlmPu_C.pdf} \hspace{0.3 cm}
\includegraphics[width=6 cm]{./Results/stat_results/stat_testing/D2umD2l_C.pdf}
\caption{(A) Schematic diagram supporting defintions of nonparametric tail dependence statistics; (B) Results of tests of $\cor_l-\cor_u$, (C) $\Ps_l-\Ps_u$, and (D) $\Dsq_u-\Dsq_l$ reveal the known asymmetry (more lower tail than upper) of tail dependence of the Clayton copula with $n=35$ data points. \label{fig_stat_image}}
\end{center}
\end{figure}

@@

<!--I deleted the phrase about unit testing because unit testing should be mentioned in the supp mat-->
We tested our statistics by testing whether $\cor_l-\cor_u$, $\Ps_l-\Ps_u$, and $\Dsq_u-\Dsq_l$ revealed the asymmetry of tail dependence known to be present in data simulated from various copulas. Fig. \ref{fig_stat_image}(B - D) represent the results for Clayton copulas with $n=35$ data points and $100$ simulations (for details see Appendix \ref{SM-Test_npa_stat}). 
<!--***DAN: There is either too little or too much here in this 
paragraph. The reader cannot get the main
idea, really, of what was done for these tests, from this paragraph alone - they will have to
look at the appendix. They need to be able to get the main idea. So you could either add detail,
or else take away detail and just say something like "We tested our statistics by testing whether 
$\cor_l-\cor_u$, $\Ps_l-\Ps_u$, and $\Dsq_u-\Dsq_l$ revealed asymmetry of tail dependence 
known to be present in data simulated from various copulas (Appendix X)" and then stop there.
Then you would remove panels B-D from Fig. 1. At the moment I prefer this second option, but 
my preference is not strong, so if you prefer to keep the material and add detail that would be 
fine.-->

For bivariate datasets, we compared values 
of the statistics $\cor_{l_b,u_b}$,
$\Ps_{l_b,u_b}$, $\Dsq_{l_b,u_b}$, $\cor_l-\cor_u$, $\Ps_l-\Ps_u$, and $\Dsq_u-\Dsq_l$ to 
distributions of values of the same statistics computed on surrogate datasets that were produced from 
the empirical data by randomizing it in a special way. The surrogate/randomized datasets had exactly 
the same marginal distributions as the empirical data and had very similar Kendall or Spearman 
correlation (the algorithm had two versions, one for preserving each correlation coefficient), 
but had Gaussian copula structure. Thus our comparisons tested the null hypothesis 
that our statistics took values on the empirical data no different from what would have been expected 
if the copula structure of the data were Gaussian, but the data were otherwise 
statistically unchanged.
Details of how surrogates were prroduced are in XYZ, and code is in ABC.
<!--***DAN: I think the below material belongs in supp mat, because the reader can get the
main idea via the above text, and only some readers will be interested in the detailed 
algorithm. Please move it there, and then make the above links XYZ and ABC into live links
to the appropriate locations in the sup mat.

@@

Because $(x_{\sigma_x(1)},\ldots,x_{\sigma_x(n)})$ 
is a permutation of $(x_1,\ldots,x_n)$ and $(y_{\sigma_y(1)},\ldots,y_{\sigma_y(n)})$ is a 
permutation 
of $(y_1,\ldots,y_n)$, the marginal distributions of the surrogate data are exactly the same as 
those of the original data. 
Because ranks of the surrogate data are matched to ranks of the $a_i$ and $b_i$, the
Kendall correlation of the surrogate dataset will be the same as that of the $a_i$ and $b_i$, and therefore 
will be close to $\tau$ (differences arising only through sampling variation). If instead of $\theta_\tau$
a parameter value $\theta_\rho$ is used such that the Spearman correlation of $\mathcal{B}(\theta_\rho)$ is the same
as that of the $x_i$ with the $y_i$, then Spearman correlations of surrogates will very nearly equal (except for 
sampling variation) the Spearman correlation of the empirical data.  
-->

>>>>

<!--**SG: I think normal copula would be better as "N" is the abbreviation for that but for "G" means gumbel not gaussian.
If you agree with me, then I will made all "gaussian" as "normal" copula in both main and supmat.-->

@@

Left- and right-tail dependence quantify the extent to which two ranked
variables $u_i$, $v_i$ for $i=1,...,n$ 
(see \nameref{Model_selection} for defintions of $u_i$ and $v_i$)
are related to each other in the left and right tails, respectively, of their 
distributions [@nelsen2007_copula, @joe2014_dependence]. 
Our nonparametric study of tail dependence uses three
statistics which can quantify the extent to which $u_i$ and $v_i$ are 
related in any part of their distributions. We here describe the 
statistics. The statistics were formulated with positively associated
variables in mind. All the variables we study are positively associated
when they are significantly associated. So definitions and associated 
diagrams will assume positive associations, though generalizations are 
possible. Given two bounds 
$0 \leq l_b < u_b \leq 1$, we define the parallel lines $u+v=2l_b$ and 
$u+v=2u_b$ which intersect the unit square in which the $u_i$ and $v_i$ 
fall (these are the blue and red lines, respectively, in 
Fig. \ref{fig_stat_image}). Our statistics quantify the dependence
between $u_i$ and $v_i$ in the region bounded by these lines.
Using $l_b=0$ and some value $u_b\leq 0.5$ can give information about 
dependence in the left parts of the distributions of $u$ and $v$, and 
using $u_b=1$ and $l_b \geq 0.5$ can give information about dependence in 
the right parts of the distributions.

\begin{figure}[!h]
\begin{center}
\includegraphics[width=6 cm,height=6 cm]{./stat_image.jpg} \hspace{0.6 cm}
\caption{Schematic diagram supporting defintions of nonparametric tail dependence statistics \label{fig_stat_image}}
\end{center}
\end{figure}

<!--
\begin{figure}[!h]
\textbf{ \hspace{6 cm} (A) \hspace{5.6 cm} (B)} \\
\vspace{-0.6 cm}
\begin{center}
\hspace{0.5 cm}
\includegraphics[width=5 cm,height=4.3 cm]{./stat_image.jpg} \hspace{0.6 cm}
\includegraphics[width=6 cm]{./Results/stat_results/stat_testing/CorlmCoru_C.pdf}
\end{center}
\textbf{ \hspace{6 cm} (C) \hspace{5.7 cm} (D)} \\
\vspace{-0.6 cm}
\begin{center}
\hspace{0.2 cm}
\includegraphics[width=6 cm]{./Results/stat_results/stat_testing/PlmPu_C.pdf} \hspace{0.3 cm}
\includegraphics[width=6 cm]{./Results/stat_results/stat_testing/D2umD2l_C.pdf}
\caption{(A) Schematic diagram supporting defintions of nonparametric tail dependence statistics; (B) Results of tests of $\cor_l-\cor_u$, (C) $\Ps_l-\Ps_u$, and (D) $\Dsq_u-\Dsq_l$ reveal the known asymmetry (more lower tail than upper) of tail dependence of the Clayton copula with $n=35$ data points. \label{fig_stat_image}}
\end{center}
\end{figure}
-->

@@

<!--I deleted the phrase about unit testing because unit testing should be mentioned in the supp mat-->
We tested our statistics by testing whether $\cor_l-\cor_u$, $\Ps_l-\Ps_u$, and $\Dsq_u-\Dsq_l$ revealed the asymmetry of tail dependence known to be present in data simulated from various copulas (for details see Appendix \ref{SM-Test_npa_stat}). 
<!--***DAN: There is either too little or too much here in this 
paragraph. The reader cannot get the main
idea, really, of what was done for these tests, from this paragraph alone - they will have to
look at the appendix. They need to be able to get the main idea. So you could either add detail,
or else take away detail and just say something like "We tested our statistics by testing whether 
$\cor_l-\cor_u$, $\Ps_l-\Ps_u$, and $\Dsq_u-\Dsq_l$ revealed asymmetry of tail dependence 
known to be present in data simulated from various copulas (Appendix X)" and then stop there.
Then you would remove panels B-D from Fig. 1. At the moment I prefer this second option, but 
my preference is not strong, so if you prefer to keep the material and add detail that would be 
fine.-->
<!--**SG: Agreed with 2nd option and did things as necessary-->

For bivariate datasets, we compared values 
of the statistics $\cor_{l_b,u_b}$,
$\Ps_{l_b,u_b}$, $\Dsq_{l_b,u_b}$, $\cor_l-\cor_u$, $\Ps_l-\Ps_u$, and $\Dsq_u-\Dsq_l$ to 
distributions of values of the same statistics computed on surrogate datasets that were produced from 
the empirical data by randomizing it in a special way. The surrogate/randomized datasets had exactly 
the same marginal distributions as the empirical data and had very similar Kendall or Spearman 
correlation (the algorithm had two versions, one for preserving each correlation coefficient), 
but had Gaussian copula structure. Thus our comparisons tested the null hypothesis 
that our statistics took values on the empirical data no different from what would have been expected 
if the copula structure of the data were Gaussian, but the data were otherwise 
statistically unchanged.
Details of how surrogates were prroduced are in Appendix \ref{SM-surrog_test}, and code is in [`copsurrog2d`](https://github.com/sghosh89/BIVAN/blob/master/copsurrog2d.R).
<!--***DAN: I think the below material belongs in supp mat, because the reader can get the
main idea via the above text, and only some readers will be interested in the detailed 
algorithm. Please move it there, and then make the above links XYZ and ABC into live links
to the appropriate locations in the sup mat.

@@

Because $(x_{\sigma_x(1)},\ldots,x_{\sigma_x(n)})$ 
is a permutation of $(x_1,\ldots,x_n)$ and $(y_{\sigma_y(1)},\ldots,y_{\sigma_y(n)})$ is a 
permutation 
of $(y_1,\ldots,y_n)$, the marginal distributions of the surrogate data are exactly the same as 
those of the original data. 
Because ranks of the surrogate data are matched to ranks of the $a_i$ and $b_i$, the
Kendall correlation of the surrogate dataset will be the same as that of the $a_i$ and $b_i$, and therefore 
will be close to $\tau$ (differences arising only through sampling variation). If instead of $\theta_\tau$
a parameter value $\theta_\rho$ is used such that the Spearman correlation of $\mathcal{B}(\theta_\rho)$ is the same
as that of the $x_i$ with the $y_i$, then Spearman correlations of surrogates will very nearly equal (except for 
sampling variation) the Spearman correlation of the empirical data.  
-->
<!--**SG: Done-->

### 2018-06-26 SupportingInformation.Rmd
trivial
adding response by **SG

<!--checkpoint chunk is moved to makefile.Rmd-->

@@

>>>>

@@

# Surrogate testing with bivariate datasets for non-parametric statistics \label{surrog_test}

For bivariate datasets, surrogates were produced as follows.

- Given data $(x_i,y_i)$ for $i=1,\ldots,n$ and a 
one-parameter target copula family $\mathcal{B}(\theta)$ (here $\theta$ denotes the parameter), let $\tau$ be the
Kendall correlation of the $x_i$ with the $y_i$ and find the parameter value $\theta_\tau$ for which 
$\mathcal{B}(\theta_\tau)$ has Kendall correlation $\tau$ (this is possible for the one-parameter copula families of this 
study using the `iTau` function of the `VineCopula` package). 
- Generate data $(a_i,b_i)$, $i=1,\ldots,n$ from 
$\mathcal{B}(\theta_\tau)$. 
- Permute the ordered set $(x_1,\ldots,x_n)$ such that the permutation
$(x_{\sigma_x(1)},\ldots,x_{\sigma_x(n)})$ has $\rank(x_{\sigma_x(i)})$ equal to $\rank(a_i)$ for all $i$, where 
$\rank(x_{\sigma_x(i)})$ is the rank of $x_{\sigma_x(i)}$ in the set $(x_{\sigma_x(1)},\ldots,x_{\sigma_x(n)})$, and
$\rank(a_i)$ is the rank of $a_i$ in the set $(a_1,\ldots,a_n)$ (here $\sigma_x$ is a permutation of the 
indices $1,\ldots,n$). 
- Likewise permute $(y_1,\ldots,y_n)$
such that the permutation
$(y_{\sigma_y(1)},\ldots,y_{\sigma_y(n)})$ has $\rank(y_{\sigma_y(i)})$ equal to $\rank(b_i)$ for all $i$ ($\sigma_y$ is 
another permutation of $1,\ldots,n$).
- The surrogate dataset is $(x_{\sigma_x(i)},y_{\sigma_y(i)})$ for $i=1,\ldots,n$.

Our code for this algorithm is [`copsurrog2d`](https://github.com/sghosh89/BIVAN/blob/master/copsurrog2d.R) in the 
[`BIVAN`](https://github.com/sghosh89/BIVAN) repository.

Because $(x_{\sigma_x(1)},\ldots,x_{\sigma_x(n)})$ 
is a permutation of $(x_1,\ldots,x_n)$ and $(y_{\sigma_y(1)},\ldots,y_{\sigma_y(n)})$ is a 
permutation 
of $(y_1,\ldots,y_n)$, the marginal distributions of the surrogate data are exactly the same as 
those of the original data. 
Because ranks of the surrogate data are matched to ranks of the $a_i$ and $b_i$, the
Kendall correlation of the surrogate dataset will be the same as that of the $a_i$ and $b_i$, and therefore 
will be close to $\tau$ (differences arising only through sampling variation). If instead of $\theta_\tau$
a parameter value $\theta_\rho$ is used such that the Spearman correlation of $\mathcal{B}(\theta_\rho)$ is the same
as that of the $x_i$ with the $y_i$, then Spearman correlations of surrogates will very nearly equal (except for 
sampling variation) the Spearman correlation of the empirical data.
